### 2015-03-26 thesis-proposal/problem.tex
updated proposal

Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
There has been substantial progress on many of the foundational semantic issues that are relevant to the design of general-purpose languages over the past  decades, much of it driven by advances in type theory. However, when more pragmatic language design issues are brought into consideration, a broadly agreeable set of primitives remains elusive, as evidenced by the diverse array of dialects that surround around all major con\-temporary languages. 
For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with sums, products and recursive types, they all introduce various syntactic and semantic extensions and tweaks of their own novel design (we discuss a number of examples below). Tools for  constructing new ``domain-specific'' language dialects also continue to proliferate. % The situation is similar across all major language lineages. 
Why are well-informed programmers and researchers so often tempted to construct  dialects of a language like ML? 
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common reason is simply that, for human {programmers}, the \emph{syntactic cost} of expressing a construct of interest using general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This gives us precisely the semantics that we need, but the syntactic cost  is not quite ideal. In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library. For example, Ur/Web is a derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. Indeed, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g. derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}. Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, which we will return to later, help lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.

@@

The problem with this practice of using language dialects as vehicles for new syntactic and semantic primitives is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
This would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two languages, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects do, however, have an indirect impact on the practice of programming because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
These major languages have thus become more expressive over time. However, this has come at a cost: they have also ballooned in size. Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects. A conservative approach to incorporating new primitives is only sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

>>>>

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Over the past several decades, researchers have developed a substantial understanding of  the theoretical issues relevant to the design of programming languages, in large part by constructing and studying the metatheoretic properties of minimal typed lambda calculi. These minimal calculi are clearly not directly suitable for real-world software development, so language designers must expose a syntax and type structure of a more elaborate design to programmers. %syntactic and semantic primitives that capture common idioms naturally and precisely.
Language designers have many viable choices at this level, as evidenced by the diverse array of dialects that continue to proliferate around all major languages. For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own novel design (we cover several examples below). 
%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
Tools for  constructing ``domain-specific'' dialects contribute further to this proliferation of language primitives. This is alarming: why is it that well-informed programmers and researchers are so often unable to express the constructs that they need in libraries, as modes of use of a small number of ``general-purpose'' language primitives?
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common  reason is simply that, for human {programmers}, the \emph{syntactic cost} of expressing a construct of interest using general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This will give  us precisely the semantics that we need, but the syntactic cost will not be ideal. In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library. For example, Ur/Web is a derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. In fact, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g. derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}. Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, which we will return to later, help lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.

@@

The reason why this practice of using dialects as vehicles for new syntactic and semantic primitives is alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
This would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two languages, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects can, however, have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
These major languages have thus become more expressive over time. However, this has come at a cost: they have also ballooned in size. Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects. 
A conservative approach to incorporating new primitives is sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
